---
title: "The skytrackr R package"
subtitle: "an optimization based approach to geolocation by light"
author: "Koen Hufkens"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The skytrackr R package}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
set.seed(0)
library(dplyr)
library(ggplot2)
library(patchwork)
library(BayesianTools)
library(skytrackr)
```

## Abstract

Geolocation by light has revolutionised animal behavioural studies since its inception, allowing for the tracking of animals using sunset and sunrise estimations. Currently, threshold methodologies use fixed threshold values to determine sunrise and sunset from diurnal light measurements, from which the location is determined (deterministic - via equations). Setting this threshold is time demanding as often not suitable for automated processing at volume due to inherent uncertainties such as contamination in the diurnal light measurements due to clouds and rapid movements around the equator. Here, I present an optimization based methodology which leverages the whole or part of the diurnal light profile and a Bayesian based optimization approach to estimate geolocations from light. The use of additional data during twilight, or a full day, and not only a threshold based sunset or sunrise allows for automation of the process. Including more data in location estimates, based upon a physical model of sky illuminance  increases accuracy and provides uncertainty measures for each estimate, allowing for post-hoc further quality control. Convenience and robustness of the method comes at the cost of required computational power compared to fast threshold based approaches.

**Keywords:** geolocation, tracking, movement ecology, flight, illuminance

## Background

Geolocation by light has been a popular animal tracking method in ornithology. It has been successfully used to track long distance migration and foraging behaviour of swifts [@kearsley_aeroecology_2022, @norevik_windassociated_2020, @hedenstrom_annual_2016, @norevik_highly_2019, @hedenstrom_flight_2019] and other migratory birds (REFERENCES), and has revolutionised our understanding of long distance movements. 

Geolocation by light is a conceptually easy approach. It relies on the shift of sunrise and sunset times to determine longitude and the length of night (or day) between sunrise or sunset to determine latitude deterministically using astronomical equations [@ekstrom_advance_2004].

However, considerable uncertainty is introduced due to the choice of the threshold in light levels used in determining sunset or sunrise times [@bridge_advances_2013, @joo_navigating_2020]. The choice of a proper threshold and solar angle is critical as it determines the daylength and therefore estimates of latitude. “Increasing mismatch between light intensity threshold and sun elevation threshold results in a decreasing accuracy in latitudinal estimation.” Livotski. Ongoing research therefore often focuses on the use of ancillary data in addition to threshold based methods (REFERENCE), or iterative methods to constrain the search space of optimal solutions better [@merkel_probabilistic_2016]. Twilight free methods have been explored by @bindoff_twilightfree_2018 which uses a hidden markov model approach to constrain parameters using the full light level data and not explicitly sunrise and sunset thresholding.

Given an unknown starting position and a reference daylength to calibrate a threshold on this threshold is often estimated by eye. To improve upon this approach leveraging the full transition profile can be used to estimate locations [@ekstrom_advance_2004]. Despite these advances in methodology few computational resources are available to estimate positions in bulk with minimal intervention.

Here, I present the {skytrackr} R package which provides a Bayesian optimization based approach using a physical sky illuminance model [@hufkens_skylight_2022, @janiczek_computer_1987] to estimate geolocation from light levels. This package and methodology provides robust and automated geolocation estimates, addressing issues regarding the choice of a threshold and sun angle parameters while increasing robustness in case of cloudy or otherwise noisy signals.

## Implementation

The {skytrackr} R package uses a physical sky illuminance model [@hufkens_skylight_2022, @janiczek_computer_1987] in combination with Bayesian location parameter estimation [@hartig_bayesiantools_2023] approach to maximise the log likelihood between observed (logger) and modelled values. Model parameters are limited to latitude, longitude and a sky condition which scales the sky illuminance response for the day. The approach improves upon the template matching as suggested by @ekstrom_advance_2004 by including a physical component to the model rather than considering a piecewise regression based approach.

```{r echo = FALSE, message = FALSE, warning = FALSE, fig.cap="Modelled and observed illuminance values in log(lux) for three days in early July 2021. Observed values are shown as light grey dots, modelled values for the exact location are shown as a full red line. Longitudinal offset data is shown as the blue full line, while latitudinal offset modelled data for the same period is shown as a full green line."}
library(skylight)

# sequence of hours
dates <- seq(ISOdate(2021,07,10), ISOdate(2021,07,12), by = "hour")

# longitude offset
location_1 <- data.frame(
  date = dates,
  longitude = 3.74,
  latitude = 51.07
) |>
  skylight()

location_2 <- data.frame(
  date = dates,
  longitude = 20,
  latitude = 51.07
) |>
  skylight()

location_3 <- data.frame(
  date = dates,
  longitude = 3.74,
  latitude = 0
) |>
  skylight()

p <- ggplot() +
  geom_point(
    data = skytrackr::cc874,
    aes(
      date_time,
      log(lux)
    ),
    colour = "grey"
  ) +
  geom_line(
    data = location_1,
    aes(
     date,
     log(total_illuminance)
    ),
    colour = "red"
  ) +
    geom_line(
    data = location_2,
    aes(
     date,
     log(total_illuminance)
    ),
    colour = "blue"
  ) +
  geom_line(
    data = location_3,
    aes(
     date,
     log(total_illuminance)
    ),
    colour = "green"
  ) +
  labs(
    title = "Total illuminance for 3 days at the start of May 2020",
    y = "log(lux)",
    x = "Date"
  ) +
  theme_bw()

plot(p)
```

By default, the DEzs MCMC sampler [@ter_braak_differential_2008] implemented in BayesianTools [@hartig_bayesiantools_2023] is used to estimate the posterior distribution of model parameters. For each location I sample the posterior distribution of the parameters (with a thinning factor of 10), and return summary values. I return the median for best estimates of latitude, longitude and sky condition. In addition, the 5-95% quantiles on the geographic position are provided. The package also returns the Gelman-Rubin-Brooks (GRB) metric [@gelman_inference_1992], which can be used to assess the convergence of the optimization. Both the quantile ranges and the GRB metric can be used in post-hoc quality control screening.

The method as proposed should perform equally well, or better than, the Ekstrom methodology using template matching [@ekstrom_advance_2004]. However, some limitations do apply. The measured light intensities should be translated to physical units in lux to be accurately compared to the physical model. Not all sensors return physical units, with some returning an arbitrary numbers due to data limitations (compression) or hardware choices. In these cases an external calibration to convert arbitrary numbers to lux is necessary. Although performance should improve around the equinoxes, performance will remain poorer than average. However, the addition of geolocation estimation uncertainties as well as GRB metrics make screening outliers easy and consistent. The methodology should compare favourably to the "Twilight-Free" method as described by [@bindoff_twilightfree_2018], as no external packages should be loaded from github.

The package and function descriptions can be found on github at: https://github.com/bluegreen-labs/skytrackr, and can be installed from source using:

```{r eval = FALSE}
if(!require(remotes)){install.packages("remotes")}
remotes::install_github("bluegreen-labs/skytrackr")
```

## Discussion

To demonstrate the functioning of the package a small demo dataset comprised of a single day of light logging near a Common swift nest box in Ghent, Belgium was included (i.e. tag cc874). I will use this data to demonstrate the features of the package. Note that when multiple dates are present all dates will be considered. The package is friendly to the use of R piped (|>) commands. Using all default settings you can just pipe the data in to the `skytrackr()` function. The returned object will be a data frame containing the best estimate (median) of the longitude and latitude as well as 5-95% quantile as sampled from the posterior parameter distribution.

```{r}
# normally you would first read in the .lux or .glf file using
# cc874 <- stk_read_lux("cc874.lux")

# load the integrated package data
print(head(skytrackr::cc874))

# skytrackr allows for piped data input
location <- cc874 |>
  skytrackr::skytrackr()

# print the optimized parameters (i.e. the locations)
print(location)
```

Note that the returned location estimates are very close to the original location of the colony (51.8 N/ 3.73 E). These parameters can be fed back into the original (underlying) algorithm to calculate the full diurnal profile. Doing so allows you to compare the modelled (fitted) data to the observed values.

```{r echo = FALSE, fig.cap="Figure showing the fitted model results (full black line) from optimized parameters (latitude, longitude, sky conditions). Default parameters only use a limited twilight set of parameters. From the observed values (coloured dots) only the 'True' values were included in the estimation of the model parameters."}
# call the underlying model
cc874$model_results <- skytrackr::log_lux(
  par = c( location$latitude, location$longitude, location$sky_conditions),
  cc874
)

ggplot(data = cc874) +
  geom_point(
    aes(
      hour,
      log(lux),
      colour = (lux > 0.32 & lux < 400)
    )
  ) +
  scale_color_discrete() +
  geom_line(
    aes(
      hour,
      model_results
    )
  ) +
  labs(
    x = "Hour",
    y = "log(lux)",
    colour = "Optimized"
  ) +
  theme_bw()
```

This visually confirms the good fit of the model data with the observed values. Note that only a subset of the data is used in model optimization (as defined by the `range` parameter). You can assess the convergence of the model using the `grb` values (i.e the Gelman-Rubin-Brooks metric, @gelman_inference_1992), which should be lower than 1.1 to indicate the convergence of the 3 chains used in the MCMC DEz algorithm. This is further corroborated by considering the upper and lower quantiles and their spread, which are narrowly defined for both latitude and longitude.

### Sensitivity analysis

As a sensitivity analysis I set values to low (effectively 0) values, for a range of 0 - 75% of the dataset. For these subsets I estimate both latitude and longitude using either the full diurnal profile or the default settings using twilight values only. In addition, the same routine was used to introduce bright moments during nighttime (250 lux).

```{r echo = FALSE}
# load sensitiviy analysis
sensitivity_default <- readRDS(system.file(package = "skytrackr","extdata/sensitivity_default.rds"))
sensitivity_full <- readRDS(system.file(package = "skytrackr","extdata/sensitivity_full.rds"))
sensitivity_night <- readRDS(system.file(package = "skytrackr","extdata/sensitivity_night.rds"))

p <- ggplot(sensitivity_default) +
  geom_point(
    aes(
      0:75,
      latitude - 51.07
    )
  ) +
  labs(
    x = "% missing data",
    y = expression(paste(Delta, ' latitude'))
  )

p2 <- ggplot(sensitivity_default) +
  geom_point(
    aes(
      0:75,
      longitude - 3.74
    )
  ) +
  labs(
    x = "% missing data",
    y = expression(paste(Delta, ' longitude'))
  )


p3 <- ggplot(sensitivity_full) +
  geom_point(
    aes(
      0:75,
      latitude - 51.07
    )
  ) +
  labs(
    x = "% missing data",
    y = expression(paste(Delta, ' latitude'))
  )

p4 <- ggplot(sensitivity_full) +
  geom_point(
    aes(
      0:75,
      longitude - 3.74
    )
  ) +
  labs(
    x = "% missing data",
    y = expression(paste(Delta, ' longitude'))
  )

p5 <- ggplot(sensitivity_night) +
  geom_point(
    aes(
      0:75,
      latitude - 51.07
    )
  ) +
  labs(
    x = "% missing data",
    y = expression(paste(Delta, ' latitude'))
  )

p6 <- ggplot(sensitivity_night) +
  geom_point(
    aes(
      0:75,
      longitude - 3.74
    )
  ) +
  labs(
    x = "% missing data",
    y = expression(paste(Delta, ' longitude'))
  )

p + p2 + p3 + p4 + p5 + p6 + 
  plot_layout(
    ncol = 2,
    guides = "auto"
  )  +
  plot_annotation(
    title = "",
    subtitle = "",
    caption = "",
    tag_levels = "a",
    tag_suffix = ")    ",
    tag_prefix = "("
  ) &
  theme_bw() +
  theme(
    panel.spacing.y = unit(0.01, "in"),
    legend.box.margin = margin(t = 5),
    legend.justification = "left",
    legend.title = element_text(
      family = "Montserrat",
      color = "grey30", size = 13
    )
  )

```

Sensitivity analysis show that especially nighttime bright moments disrupt the positional accuracy of geolocation estimates. While using performance declines steadily when considering the full profile at higher noise levels (c-d).

### Limitations

The package does not mitigate uncertainty introduced by occlusion of the sensor by feathers, or birds roosting in secluded places. However, sensitivity analysis have shown that these factors might have less effect than nighttime illumination. Regardless, performance of the optimization method remains functionally optimal in species such as swifts, which continuously fly during their non-breeding season []. Data assimilation by using pressure measurements at stationary moments (REFERENCE) could be implemented but is currently not. Further improvements can be implemented by masking of locations during optimization (based on step distance and or geographic location, such as a land water mask), and are currently not considered.

## Conclusion

The {skytrackr} R package provides an optimization based methodology which leverages the whole or part of the diurnal light profile and a Bayesian based optimization approach to estimate geolocations from light. The use of additional data during twilight, or a full day, allows for automation of the process over threshold based methods. Including more data in location estimates, based upon a physical model of sky illuminance, increases accuracy. The Bayesian framework also provides uncertainty measures for each estimate, allowing for post-hoc quality control.

## Availability and requirements

 - **Project name:** skytrackr
 - **Project homepage:** https://github.com/bluegreen-labs/skytrackr
 - **Operating system(s):** Platform independent
 - **Programming Language:** R
 - **License:** AGPLv3

## References


